---
title: "Metrics"
output: html_document
---


## Preliminary steps

Load packages

```{r}
library(tidyverse)
```

Connect to the database

```{r}
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = "../data/faers_22q2.duckdb", read_only = TRUE)
```


```{r}
dbplyr::src_dbi(con)
```

Set up table references

```{r}
demo <- tbl(con, "demo")
drug <- tbl(con, "drug")
indi <- tbl(con, "indi")
outc <- tbl(con, "outc")
reac <- tbl(con, "reac")
rpsr <- tbl(con, "rpsr")
ther <- tbl(con, "ther")
```

## Analysis



Get drug-reaction pairs that occur more than 10 times.

```{r}
pairs <- reac %>%
  select(pt, primaryid) %>%
  distinct() %>%
  inner_join(
    drug %>% select(drugname, primaryid) %>% distinct(),
    by = "primaryid"
  ) %>%
  group_by(drugname, pt) %>%
  summarise(count = n()) %>%
  filter(count > 10)

pairs %>%
  arrange(desc(count))
```


Get counts of drugs and reactions.

```{r}
drug_totals <- drug %>%
  select(drugname, primaryid) %>%
  distinct() %>%
  group_by(drugname) %>%
  summarise(drug_total = n())
```

```{r}
reaction_totals <- reac %>%
  select(pt, primaryid) %>%
  distinct() %>%
  group_by(pt) %>%
  summarise(reaction_total = n())
```

Compute the total number of observations.

```{r}
total_obs <- reac %>%
  select(primaryid) %>%
  distinct() %>%
  summarise(count = n()) %>%
  collect()

total_obs <- total_obs$count[1]
total_obs
```

Now create the contingency table.

```{r}
contingency_tables <- pairs %>%
  inner_join(drug_totals, by = "drugname") %>%
  inner_join(reaction_totals, by = "pt") %>%
  mutate(
    a = as.numeric(count),
    b = as.numeric(drug_total - count) + 1,
    c = as.numeric(reaction_total - count) + 1,
    d = as.numeric(total_obs - reaction_total - drug_total + count)
  )
```


Now compute the proportional reporting ratio, reporting odds ratio, and the observed over expected.

```{r}
metrics <- contingency_tables %>%
  mutate(
    prr = (a / (a + b)) / (c / (c + d)),
    ror = (a / b) / (c / d),
    ooe = (a / (a + b)) / ((a + c) / (a + b + c + d))
  )
```

```{r}
metrics %>%
  select(drugname, pt, count, prr, ror, ooe) %>%
  arrange(desc(count))
```

```{r}
metrics %>%
  select(drugname, pt, count, prr, ror, ooe) %>%
  arrange(desc(ror))
```


## Wrapping up

Disconnect from the database

```{r}
DBI::dbDisconnect(con, shutdown = TRUE)
```