---
title: "Zantac"
output: html_document
---


In this notebook, I do some exploration of the most frequently occurring drug in the dataset: Zantac.



## Preliminary steps


Load packages

```{r}
library(tidyverse)
library(DBI)
```



Connect to the database

```{r}
con <- dbConnect(duckdb::duckdb(), dbdir = "../data/faers_22q2.duckdb", read_only = TRUE)
```

## Analysis


Which events had the most medications involved?

```{r}
dbGetQuery(
  con,
  "SELECT
    primaryid,
    COUNT(*) AS num_medications
  FROM drug
  GROUP BY primaryid
  ORDER BY COUNT(*) DESC"
) %>%
  as_tibble()
```



Which are the drugs that have the most reports?

```{r}
dbGetQuery(
  con,
  "SELECT
    drugname,
    COUNT(DISTINCT primaryid) AS num_events
  FROM drug
  GROUP BY drugname
  ORDER BY COUNT(DISTINCT primaryid) DESC"
) %>%
  as_tibble()
```

The drug involved in the most reports is Zantac, and it is the highest by more than a factor of 2.



Take a look at the reac table.

```{r}
dbGetQuery(
  con,
  "SELECT * FROM reac"
) %>%
  as_tibble()
```




Further examine Zantac, the drug that had the most reports in Q2 2022.

```{r}
zantac_reactions <- dbGetQuery(
  con,
  "SELECT
    reac.pt,
    COUNT(DISTINCT drug.primaryid) AS num_reactions,
    COUNT(DISTINCT drug.primaryid) * 1.0 / (SELECT COUNT(DISTINCT primaryid) FROM drug WHERE drugname = 'ZANTAC') AS share_of_total_reactions
  FROM drug INNER JOIN reac ON (drug.primaryid = reac.primaryid)
  WHERE drug.drugname = 'ZANTAC'
  GROUP BY reac.pt
  ORDER BY COUNT(DISTINCT drug.primaryid) DESC"
) %>%
  as_tibble()

zantac_reactions
```

All of the top 10 highest reported events are cancer. Prostate cancer comprises roughly 13% of all Zantac reports.

Let's look at the rest of the data.

```{r}
all_reactions <- dbGetQuery(
  con,
  "SELECT
    pt,
    COUNT(DISTINCT primaryid) AS num_reactions,
    COUNT(DISTINCT primaryid) * 1.0 / (SELECT COUNT(DISTINCT primaryid) FROM reac) AS share_of_total_reactions
  FROM reac
  GROUP BY pt
  ORDER BY COUNT(*) DESC"
) %>%
  as_tibble()

all_reactions %>%
  print(n = 40)
```

Prostate cancer is the 40th most common occurence, with 5241. For 4762 of these, Zantac must have been a reported medication.

```{r}
zantac_shares <- zantac_reactions %>%
  slice_head(n = 40) %>%
  rename(num_zantac_reactions = num_reactions) %>%
  left_join(all_reactions, by = "pt") %>%
  select(!starts_with("share")) %>%
  mutate(zantac_share = num_zantac_reactions / num_reactions)

zantac_shares
```



```{r}
zantac_shares %>%
  ggplot(mapping = aes(zantac_share, fct_reorder(pt, zantac_share))) +
  geom_col(width = 0.8)
```

For many of these events, Zantac was reported as a medication in the vast majority of reports.



What drugs was Zantac prescribed with?

```{r}
zantac_bystanders <- dbGetQuery(
  con,
  "SELECT
    drugname,
    COUNT(DISTINCT primaryid) AS num_occurrences,
    COUNT(DISTINCT primaryid) * 1.0 / (SELECT COUNT(DISTINCT primaryid) FROM drug WHERE drugname = 'ZANTAC') AS occurrence_rate
  FROM drug
  WHERE
    NOT drugname = 'ZANTAC'
    AND primaryid IN (SELECT primaryid FROM drug WHERE drugname = 'ZANTAC')
  GROUP BY drugname
  ORDER BY COUNT(DISTINCT primaryid) DESC"
) %>%
  as_tibble()

zantac_bystanders
```




How often was Zantac prescribed alone?

```{r}
zantac_bystanders <- dbGetQuery(
  con,
  "SELECT
    'ZANTAC ALONE' AS drugname,
    COUNT(DISTINCT primaryid) AS num_occurrences,
    COUNT(DISTINCT primaryid) * 1.0 / (SELECT COUNT(DISTINCT primaryid) FROM drug WHERE drugname = 'ZANTAC') AS occurrence_rate
  FROM drug
  WHERE
    drugname = 'ZANTAC'
    AND primaryid NOT IN (SELECT primaryid FROM drug WHERE NOT drugname = 'ZANTAC')"
) %>%
  as_tibble() %>%
  add_row(zantac_bystanders)
```

```{r}
zantac_bystanders %>%
  slice_head(n = 4) %>%
  ggplot(mapping = aes(occurrence_rate, fct_reorder(drugname, occurrence_rate))) +
  geom_col(width = 0.8)
```

It is almost always prescribed alone.
The drug that it is most frequently prescribed with is Ranitidine, and it happens roughly 3% of the time in this dataset.

## Wrapping up

Disconnect from the database

```{r}
dbDisconnect(con, shutdown = TRUE)
```